---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# excelDataGuide

<!-- badges: start -->
<!-- badges: end -->

**excelDataGuide** is an R package designed to streamline the process of importing data from spreadsheet *data reporting templates* (DRTs) into R.

A *data reporting template* is a standardized spreadsheet file (in either xls or xlsx format) used for reporting and processing experimental data. These templates significantly reduce the time required for data analysis and encourage users to present their data in a structured format, minimizing errors and misinterpretations.

The **excelDataGuide** package eliminates the need for data analysts to write and maintain complex code for reading data from various complex spreadsheet DRTs. Additionally, it offers a robust framework for validating data, ensuring that the correct data types are utilized, and facilitating data wrangling when necessary. This functionality supports *Interoperability* for DRTs, a key aspect of the [FAIR](https://www.go-fair.org/fair-principles/) principles.

The package features a user-friendly interface for extracting data from Excel files and converting it into R objects. It accommodates three types of data structures: key-value pairs, tabular data, and microplate-formatted data. The locations of these structures within the Excel template are specified by a **data guide**, which is a YAML file — a structured format that is both human- and machine-readable.

## Installation

You can install the development version of excelDataGuide in a recent version of R from GitHub with:

``` r
# install.packages("pak")
pak::pak("SystemsBioinformatics/excelDataGuide")
```

## Usage

The basic usage of the package requires only one command with two file paths: the path to the Excel data file and the path to the data guide file. Here is an example:

```{r example}
library(excelDataGuide)
datafile <- system.file("extdata", "example_data.xlsx", package = "excelDataGuide")
guidefile <- system.file("extdata", "example_guide.yml", package = "excelDataGuide")
data <- read_data(datafile, guidefile)
```

The output of the `read_data()` function is a list object the format of which is determined for a large part by the design of the data guide.

## How it works

When you design a template spreadsheet file for data reporting and analysis you also create a *data guide* file that specifies the structure and location of the data in the template. Examples of a template with data and of a data guide are provided in the package (`system.file("extdata", "example_data.xlsx", package = "excelDataGuide")` and `system.file("extdata", "example_guide.yml", package = "excelDataGuide")`).

Once you have entered the data and metadata in a template you can use the `read_data()` function in the package to extract the data into R with a single command. The package will check and coerce the data types to the required formats.

Details about the data guide format and how to write one as well as about how to design a template can be found in the package vignettes.

## Future work

- Complete the vignette ([issue](https://github.com/SystemsBioinformatics/excelDataGuide/issues/2))
- Provide guide and template structures for data types without upper size limit, typically time series with no pre-determined length ([issue](https://github.com/SystemsBioinformatics/excelDataGuide/issues/1)).
