---
title: "Excel Data Guide to R"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Excel_DRT_Guide_to_R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, echo=FALSE}
library(excelDataGuide)
```

The goal is to have both spreadsheet and scripts as data analysis tools. 

Importantly, **the source of all data is the spreadsheet.** This concerns 
metadata, parameters like acceptance criteria, concentrations and measured data.
This guarantees that calculation in the spreadsheet and in the scripts are based
on the same underlying data.

Concerning calculated data it may or may not be useful to let the spreadsheet be
the source of such data for the script as well. This may be particularly useful 
when it concerns calculations that are carried out automatically upon entry
of data by the user. 

Part of these data, like acceptance criteria, is determined in the SOP and fixed
in the spreadsheet template, whereas other data may vary per experiment and is
entered by the user.

To facilitate automatic reading from the spreadsheet by scripts data must be 
in either of these three formats:

- **keyvalue** format. Here, the key and value are placed in horizontally 
adjacent cells. The key is to be used as the parameter name in the
scripts and should conform to variable naming rules for the scripting language
used. The key is found in th left-most cell of a cell range. The value can be a
single value (one cell) or a vector of values (multiple cells).
- **platedata** format. Here the data are registered in the same row-column 
format as the microplate in which the experiment was performed. The first row
contains the variable name in its left-most cell, and is followed by (integer) 
column names. Every subsequent row contains the row name (in capital letters) 
followed by the values for each well. Both variable name and data are read by 
the script. The column and row names are ignored. Therefore, the first row 
and column in the range could also be empty, except for the variable name. 
- **table** format. This is the format for tabular data where columns represent
variables and rows represent items in which these variables are assessed. Column
names are written in the first row.

The keyvalue format will be mostly used for metadata and parameters. All keyvalue 
will be aggregated in a single named list.

The platedata format will be used for measured data and data concerning 
concentrations in the plate wells. All ranges will be aggregated in a single 
data frame with reported variables as column names, including the column names 
"row" and "col", corresponding to the row and column names of the plate.

Clearly, to make sure that calculations made in the spreadsheet and in the 
script use the same values, the spreadsheet should use parameter values *etc.*
by the (preferably absolute) cell-reference mechanism, whereas the script should
use these values by reference to their variable names.

Every spreadsheet template should be accompanied by a guide indicating the 
sheets and ranges in which keyvalue and platedata formatted data are to be found
in the filled out template. This guide is a yaml file.

This guide is structured as follows:

```{yaml}
template.name: competition
template.version: "9.0"
plate.format: 96
locations:
  - sheet: _parameters
    type: keyvalue
    translate: false
    varname: metadata
    ranges: 
      - A2:B4
      - A6:B7
      - A9:D9
  - sheet: _data
    type: platedata
    translate: false
    varname: platedata
    ranges: 
      - A2:K15
  - sheet: metadata
    type: keyvalue
    translate: true
    varname: metadata
    ranges:
      - A1:B14
      - A16:B20
  - sheet: _data
    type: table
    translate: false
    varname: rejections
    ranges:
      - A101:B111
translations:
  long name 1: shortname1
  long name 2: shortname2
```

A guide must contain the following elements:

- **template.name**: the name of the data reporting template.
- **template.version**: the version of the data reporting template.
- **plate.format**: the format of the microplate used in the experiment (valid values are 24, 48, 96 and 384).
- **locations**: a list of locations in the spreadsheet where data are to be found. Each location is a list of elements.
- **translations**: a list of translations between long and short names for variables.

The location data indicate where data are to be found, whereas the translation 
part contains translations between long and short names for variables. Short 
names are used as variable names in the scripts, whereas long names may be used 
in the spreadsheet, in particular when these are visible to the user. In that 
case the names should be translated before using them in the script. Reverse 
translations may be used by the script in the output document.

Required elements in a location are:

- **sheet**: the name of the sheet in which the data are to be found.
- **type**: the format of the data in the range.
- **translate**: (*true*, *false*) whether the variable names should be translated before use in the script.
- **varname**: the name of the variable in which the data will be available in the script.
- **ranges**: the ranges in which the data are to be found.

Furthermore, an optional element **atomicclass** can be provided which can have
values "character", "numeric", "integer" or "logical". By default, values are
converted to character, but if desired otherwise as indicated by the 
**atomicclass** element then values are coerces. Note that coercion is performed
by the functions `as.character`, `as.numeric`, `as.integer` and `as.logical` 
respectively.

The version should correspond to that reported in the template itself, 
otherwise the file or the template is invalid. The user of this package 
should take care of this check.

