---
title: "Writing data guides"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{writing_data_guides}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = FALSE
)
```

```{r setup}
library(excelDataGuide)
```

# Writing a guide

Every spreadsheet template should be accompanied by a data guide — a YAML file
that documents where data are located within the template and how they should be
interpreted. This guide serves as the blueprint that tells the `excelDataGuide`
package exactly where to find and how to read your data.

A data guide is a human-editable, computer-readable YAML file. YAML syntax is
simple: use colons (`:`) for key-value pairs, hyphens (`-`) for lists, and
indentation for nesting. This section walks you through constructing a guide
step by step.

## Step 1: Define the header

Every guide begins with five pieces of metadata that describe the template itself:

``` yaml
guide.version: '1.0'
template.name: myassay
template.min.version: '1.0'
template.max.version: '2.0'
plate.format: 96
```

- **`guide.version`** (*required*): The version of the guide itself (not the template).
  Use `major.minor` format, e.g., `'1.0'` or `'2.1'`. This allows you to track
  changes to the guide independently from the template.

- **`template.name`** (*required*): A short, descriptive name for the template
  (e.g., `'myassay'`, `'fluorescence_assay'`). This is used to identify which
  template a guide belongs to and is checked at runtime if you enable template
  name validation.

- **`template.min.version`** (*required*): The minimum template version for which
  this guide is compatible. Use `major.minor` format (e.g., `'1.0'`).

- **`template.max.version`** (*required*, can be null): The maximum template version
  for which this guide is compatible. If there is no upper limit, use `~` (YAML
  null) to indicate "any newer version is OK". For example, `template.max.version: ~`
  means the guide works with all template versions from `template.min.version`
  onwards.

- **`plate.format`** (*conditionally required*): The microplate format used in your
  experiments. Valid values are `24`, `48`, `96`, or `384` (referring to the
  number of wells). This is **required only if your guide includes any
  `platedata` locations** (see below). The guide uses this to validate that
  your platedata ranges have the correct dimensions.

## Step 2: Add your first location — template version cell

The `.template` location is a special, **required** location that records the
template version from a single cell. This allows the package to check at runtime
that the data file was created with a compatible template version.

``` yaml
locations:
  - sheet: description
    type: cells
    varname: .template
    translate: false
    variables:
      - name: version
        cell: B2
```

**What this does:**
- Reads cell `B2` from the sheet named `description`
- Stores its value in a variable called `version` within the `.template` group
- `translate: false` means no translation is applied (the cell value is used as-is)

This location **must be present in every guide**. The package uses the version
value to check compatibility with `template.min.version` and `template.max.version`.

## Step 3: Add metadata as key-value pairs

Metadata such as experimenter names, dates, and study IDs are typically stored as
key-value pairs in your template. One location entry can index multiple ranges.

``` yaml
  - sheet: description
    type: keyvalue
    varname: metadata
    translate: true
    atomicclass:
      - character
      - character
      - date
      - numeric
    ranges:
      - A5:B8
      - A10:B12
```

**What this does:**
- Reads two ranges (`A5:B8` and `A10:B12`) from the `description` sheet
- The left column of each range contains keys (variable names); the right column contains values
- `translate: true` means the keys will be translated from long names (in the
  spreadsheet) to short names (used in your script) using the `translations`
  section defined later
- `atomicclass` is a list of data types, one per key-value pair in order. In this
  example, the first three key-value pairs should be coerced to character, character,
  and date respectively, then numeric for any additional pairs (coercion cycles
  through the list if needed). All values are coerced to character by default if
  `atomicclass` is omitted.

**Combining multiple ranges into one variable:** Multiple ranges are combined into
a single named list (`metadata` in this example). This is convenient when metadata
is scattered across multiple sheets or ranges but logically belongs together.

## Step 4: Add experiment parameters

Parameters that are constant across all experiments (e.g., acceptance criteria,
standard concentrations defined in SOPs) are best stored in a dedicated `_parameters`
sheet. They are indexed like metadata:

``` yaml
  - sheet: _parameters
    type: keyvalue
    varname: parameters
    translate: false
    atomicclass: numeric
    ranges:
      - A2:B5
      - A8:B10
```

Here, all values will be coerced to numeric. Use `translate: false` if your
parameter names are already in the format you want in the script.

## Step 5: Add measured plate data

If your template records plate data (e.g., fluorescence measurements from a 96-well
plate), use the `platedata` type:

``` yaml
  - sheet: _data
    type: platedata
    varname: plate
    translate: false
    atomicclass:
      - character
      - numeric
      - numeric
    ranges:
      - A1:M9
      - A11:M19
```

**What this does:**
- Reads data in plate format (e.g., well positions and measurements)
- Each range is a block of data in the same row-column layout as the physical plate
- Multiple ranges are stacked vertically into a single data frame
- The resulting data frame includes columns `row` and `col` (well positions) plus
  one column per measured variable, each coerced to its corresponding `atomicclass`

In this example, the first three variables will be character, numeric, numeric
respectively. The first range (`A1:M9`) represents rows A–H of the plate (8 rows)
with 12 columns of data.

## Step 6: Add tabular data

Tables with column headers and multiple rows of data are indexed using the `table`
type:

``` yaml
  - sheet: results
    type: table
    varname: userresults
    translate: false
    atomicclass: numeric
    ranges:
      - C3:E10
```

**What this does:**
- Reads a table from range `C3:E10`
- The first row contains column names (variable names)
- All values are coerced to the specified `atomicclass`
- The result is a data frame, not a list

## Step 7: Add single-cell values

Sometimes you need to read isolated values from single cells and give them names
in the guide (rather than relying on names in the spreadsheet):

``` yaml
  - sheet: analysis
    type: cells
    varname: qc_checks
    translate: false
    atomicclass: numeric
    variables:
      - name: spread_well_1
        cell: G6
      - name: spread_well_2
        cell: G33
```

**What this does:**
- Reads individual cells and assigns them names specified in the guide
- `spread_well_1` gets the value from cell `G6`, `spread_well_2` from `G33`
- All values are coerced to numeric
- The result is a named list

This is useful when values are scattered or when the spreadsheet's cell-naming
scheme doesn't align with your script's variable-naming scheme.

## Step 8: Add translations

If your spreadsheet uses long, user-friendly variable names (e.g., "Date of experiment")
but your script prefers short, code-friendly names (e.g., `date`), define a
`translations` section:

``` yaml
translations:
  - long: "Date of experiment"
    short: date
  - long: "Experimenter name"
    short: experimenter
  - long: "Study identifier"
    short: studyID
  - long: "Plate identifier"
    short: plateID
```

**When to use translations:**
- Set `translate: true` in a location to apply these translations to its variable
  names
- Translations are optional; if you don't define any, set `translate: false` in
  all locations
- The package provides `long_to_shortnames()` and `short_to_longnames()` functions
  to convert between formats

## Complete example guide

Below is a complete, working guide for a small fluorescence assay template:

``` yaml
guide.version: '1.0'
template.name: simple_fluorescence
template.min.version: '1.0'
template.max.version: ~
plate.format: 96

locations:
  # Required: template version
  - sheet: description
    type: cells
    varname: .template
    translate: false
    variables:
      - name: version
        cell: B2

  # Metadata: user and experiment information
  - sheet: description
    type: keyvalue
    varname: metadata
    translate: true
    atomicclass:
      - character
      - character
      - date
      - numeric
    ranges:
      - A5:B8

  # Parameters: SOP-defined constants
  - sheet: _parameters
    type: keyvalue
    varname: parameters
    translate: false
    atomicclass: numeric
    ranges:
      - A2:B5

  # Plate data: fluorescence measurements
  - sheet: _data
    type: platedata
    varname: fluorescence
    translate: false
    atomicclass: numeric
    ranges:
      - A1:M9
      - A11:M19

  # Quality control: derived metrics from analysis sheet
  - sheet: analysis
    type: cells
    varname: qc
    translate: false
    atomicclass: numeric
    variables:
      - name: plate_uniformity
        cell: D3
      - name: signal_range
        cell: D4

translations:
  - long: "Experimenter name"
    short: experimenter
  - long: "Date of experiment"
    short: date
  - long: "Temperature (°C)"
    short: temperature
  - long: "Sample concentration (µM)"
    short: concentration
```

## Validation with CUE

Before using your guide in your scripts, validate it against the CUE schema to
catch syntax and structure errors early:

``` bash
# From the data-raw folder where the schema is located:
cd data-raw
./validate_and_sign.sh path/to/your/guide.yml
```

This command will:
1. Check your guide against `excelguide_schema.cue`
2. Abort with a detailed error message if validation fails
3. Embed a SHA256 hash (`cue.verified` field) if validation succeeds

The hash acts as proof that the guide was validated; you can verify it later
with:

``` bash
./verify_guide.sh path/to/your/guide.yml
```

In R, you can also check the hash when loading a guide:

``` r
guide <- read_guide("path/to/guide.yml", verify_hash = TRUE)
```

This will warn if the hash is missing or abort if the hash is invalid.
